---
title: "Read SAS directory"
author: "Anthony Staines"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

GERONTE <https://geronteproject.eu/> is a Research and innovation project funded by the European Union within the framework of the H2020 Research and Innovation programme - Transforming Patient Care under grant agreement No. 945218.


```{r setup, include=FALSE}
rm(list = ls())
library(stringr)
library(tidyverse)
library(haven)
library(sjlabelled)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic logic

Go to a directory.
Each contains six files

e.g.

```
STUDY_NAME_SAS_Category.txt  STUDY_NAME_SAS_Label.txt
STUDY_NAME_SAS_Data.txt      STUDY_NAME_SAS_QLU.txt
STUDY_NAME_SAS_Format.txt    STUDY_NAME_SAS_Type.txt
```

The key is the part of the name after SAS_ and before the .txt.
Please ensure that all these files exist, and have something in them!

```{r Basic lists of files and directories etc}
Directories <- list.files(path=".")
    Directories <-  # Exclude most files
      Directories[
        !str_detect(Directories, '\\.')]
## APN
APN <- list.files(path = Directories[[1]],
                  recursive = TRUE)

##EOS
EOS <- list.files(path = Directories[[2]],
                  recursive = TRUE)

##etc...

## File types - there are six.
Type <- "_([A-z]+)\\."

Types <- tibble(Type = str_extract(APN, Type)) %>%# The same in every directory
  mutate(Type = str_replace(Type,"_SAS_", "")) %>%
      mutate(Type = str_replace(Type, "\\.", "")) %>%
  mutate(Valid = FALSE)
str(Types)

rm(Type)
```

# Functions

In the labels provded, the argument are in the wrong order, so need to be switched.

```{r SWAP function}

# Need to swap 1='Yes' to 'Yes'=1

Swap_over_equal_sign <- function(Dataframe, Text) {
  
  Text = Dataframe$Text

  result <- str_match(Text, "(^[\\w]+)(=)(.+)")
  # Three matching groups
  # (1) A word from the beginning
  # (2) A literal = sign
  # (3) The rest of the line to the end

  results <- tibble(V1 = result[,1],
                    V2 = result[,2],
                    V3 = result[,3],
                    V4 = result[,4])
  
  Results <- results %>%
    mutate(V5 = # Put the terminal ; in a separate field
             ifelse(str_detect(V4,';'),';','')) %>%
    mutate(V4 = # Lose the original ;
             str_replace(V4, ';', '')) %>%
    mutate(Swapped = # Glue the pieces together backwards
             paste0(V4,V3,V2)) %>%
    select(Swapped, V1, V2, V3, V4, V5) # Reorder as desired

return(Results)
}
```

One data format (with seperators of tabs and | characters,) is most conveniently managed by str_split, which produces a list of lists as its most natural output - this function puts that into a tibble. There may, or may not, be smarter ways to do this e.g. rbindlist from data.table.

```{r List_to_tibble}

#
# From https://stackoverflow.com/questions/42642266/turn-a-list-of-lists-with-unnamed-entries-into-a-data-frame-or-a-tibble
# # Credit to A5C1D2H2I1M1N2O1R2T1
#

List_to_tibble <- function(inlist, NAMES, PRINT = FALSE) {

  stopifnot (is.logical(NAMES),
             is.logical(PRINT))

  if (NAMES){
    Names <- unlist(inlist[1]) # First row as a character variable
    inlist <- inlist[-1]
  }

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("List of lists \n")
    print(str(inlist))
    print("\n\n===========================================\n\n")}

  x <- as.data.frame(do.call(rbind, inlist))

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("as.data.frame(do.call(rbind, inlist)) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}

  x[] <- lapply(x, unlist)

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print(" x[] <- lapply(x, unlist) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}
  
  if (NAMES) {names(x) <- Names}
  
  x <- tibble(x)

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("x <- tibble(x) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}

  return(x)
}

```

Some files are empty - come up as 2B long, so exclude these.

```{r Function FILE_Test}
# FILE is a filename and relative path
# Returns a list of 2 RR[[1]] - contents of file RR[[2]] Flag is file is under 10B in size
FILE_Test <- function (FILE) {
    Test <- file.size(FILE)
    FLAG=TRUE
    if (Test > 10){ # This is 10 bytes
      RR <- readLines(paste0(FILE),
                      encoding = 'UTF-8', skipNul = TRUE)
      RR <- str_replace_all(RR, "\xff\xfe", "")
      str(RR)
      return(list(RR, FLAG))
} else {
      print(paste0('File size of the file "', FILE,
                   '" is only ', Test,
                   ' which is less than 10B, so please check it. It will not be processed further'))
        RR = ""
        FLAG = FALSE
      return(list(RR, FLAG))
  }
}

```

Work through this one subdirectory and one file at a time

```{r Set the Directory to work on}
## Remove the initial # from the relevant line to pick the directory you want to look at.

#DIRECTORY = "APN"     #1
#DIRECTORY = "EOS"     #2
#DIRECTORY = "FPROG"   #3 # BLANK DIRECTORY
#DIRECTORY = "HOSPITA" #4
#DIRECTORY = "INC"     #5
#DIRECTORY = "INSTI"   #6
#DIRECTORY = "SAE"     #7
#DIRECTORY = "TRT"     #8 # SAS FORMAT file is empty
DIRECTORY = "VIS_3M" #9
#DIRECTORY = "VIS_6M" #10
#DIRECTORY = "VIS_9M" #11
#DIRECTORY = "VIS_12M" #12
```


# SAS_Category

These look like this one, and hold value labels.

Value APN_C 1='Yes'
            0='No';
Value TYP1APN 1='Face to face'
              2='Web based'
              3='Phone call';
Value CHA1ACT 1='Yes'
              0='No';

To read these
Read the whole lot in, ignoring carriage returns, so separate only at ; 
Swap equals so 1='Yes' becomes 'Yes'=1 etc...

```{r Read SAS_Category}
ITEM = 'Category'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_Category}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
  Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(Value = # Create a logical Value to identify lines with variable names
             str_detect(Text, "^Value ")) %>% # The text 'Value ' at the start of a line
    mutate(Text = # Remove the initial Value from those lines
             str_replace(Text, "^Value ", "")) %>%
    mutate(Name = # Create a column with the variable name
             str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the [] bracket
    fill(Name) %>% # Name is only filled for the first code in each variable, so fix this.
    mutate(Text = #Remove the variable name from the Text
             str_replace(Text, "^[\\w]+", "")) %>%
    mutate(Text = str_trim(Text)) # Remove whitespace

  Results <- Input %>% Swap_over_equal_sign(Text) # Text is field with labels in it.

Output <- cbind(Input , Results)

#Source :- https://stackoverflow.com/questions/47516904/use-function-argument-as-name-for-new-data-frame-in-r

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Output,
       envir=.GlobalEnv)

#rm(RR, Input, Results, Output)
}

```

# SAS_Format

This contains formats for date variables only, for some reason?
Looks like this - 
DT_1APN ddmmyy10.
DT_2APN ddmmyy10.
DT_3APN ddmmyy10.
DT_4APN ddmmyy10.
DT_5APN ddmmyy10.

```{r Read SAS_Format}
ITEM = 'Format'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_Format}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
    Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(Name = # Create a column with the variable name
             str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
  mutate(Text = #Remove the variable name from the Text
             str_replace(Text, "^[\\w]+", "")) %>%
    mutate(Text = str_trim(Text)) # Remove whitespace

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Input %>%
         rename(Format = Text) %>%
         select(ID, Name, Format),
       envir=.GlobalEnv)

rm(RR, Input)
}

```

# SAS_Label

These are variable labels, and look like
PT_CEN="Patient Center derivé"
PT_NUM="Patient Number derivé"
PT_LET="Patient Letter derivé"
APN_C="APN"
DT_1APN="Date APN 1"
TYP1APN="Type APN 1"
CHA1ACT="Interv 1 : Treatment"
CHA1NB="Treatment 1 : Number"
GEN1ACT="Interv 1 : Lifestyle"
GEN1NB="Lifestyle 1 : Number"
CAL1ACT="Interv 1 :Extra HPC"
CAL1NB="Extra HPC 1 : Number"
HPC1ACT="Interv 1 : HPC"
HPC1NB="HPC 1 : Number"
HOS1ACT="Interv 1 : Hospit"

```{r Read SAS_Label}
ITEM = 'Label'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_Label}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
    Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(Name = # Create a column with the variable name
             str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
    mutate(Text = #Remove the variable name and the equals sign from the Text
             str_replace(Text, "^[\\w]+=", "")) %>%
    mutate(Text = # Remove the initial an final " from the SAS types
             str_replace_all(Text, "\"", "")) %>%
    mutate(Text = str_trim(Text)) # Remove whitespace

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Input %>%
         rename(Label = Text) %>%
         select(ID, Name, Label),
       envir=.GlobalEnv)


rm(RR, Input)
}


```

# SAS QLU

Looks like this 

Visit/Form/Question	SASCode	Description	Type
APN/APN_CON/PT_CEN	PT_CEN	Patient Center derivé	Integer number
APN/APN_CON/PT_NUM	PT_NUM	Patient Number derivé	Integer number
APN/APN_CON/PT_LET	PT_LET	Patient Letter derivé	Text

```{r Read SAS_QLU}
ITEM = 'QLU'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_QLU}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
    Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(Text = str_replace(Text, 'Date/Time', 'Date_Time'))  %>%
    mutate(Text = str_trim(Text)) %>%# Remove whitespace
    mutate(List_of_Lists =
             str_split(Text, pattern = "(/|\\t)"))

  QLU <- List_to_tibble(Input$List_of_Lists, NAMES = TRUE, PRINT=FALSE)
  QLU <- QLU %>%
      mutate(Type = str_replace(Type, 'Date_Time', 'Date/Time')) %>%
      mutate(Description = str_trim(Description)) # A small number of these have stray spaces at the end.

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Input %>%
         slice_head(n = -1) %>% # Remove 'Names' row
         cbind(QLU) %>%
         rename(QLU = Text),
       envir=.GlobalEnv)

rm(RR, QLU, Input)
}
```

# SAS_Type
[First seven variables are different in some way]
Looks like this
Trial :$13.
Site :$8.
Label :$50.
PersonId :5.
VisCycle :5.
FrmCycle :5.
RepeatNo :5.
PT_CEN :2.
PT_NUM :3.
PT_LET :$3.
APN_C :2.
DT_1APN :ddmmyy10.

```{r Read SAS_Type}
ITEM = 'Type'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_Type}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
    Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(Name = # Create a column with the variable name
             str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
    mutate(Text = #Remove the variable name from the Text
             str_replace(Text, "^[\\w]+", "")) %>%
    mutate(Text = str_trim(Text)) %>% # Remove whitespace
    mutate(Text = # Remove the initial : from the SAS types
             str_replace(Text, "^:", "")) %>%
    mutate(Text = str_trim(Text)) # Remove whitespace

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Input %>%
         rename(Type = Text) %>%
         select(ID, Name, Type),
       envir=.GlobalEnv)

rm(RR, Input)
}
```

# SAS_Data

```{r Read SAS_Data}
ITEM = 'Data'
FILE = paste0(DIRECTORY, "/GERONTE_FRONE20230407_SAS_", ITEM, ".txt")
RR <- FILE_Test(FILE)
Types <- Types %>% rows_update(tibble(Type = ITEM, Valid = RR[[2]]), by = "Type")
```

```{r Process SAS_Data}
if (unlist(Types %>% filter (Type == ITEM) %>% select(Valid))){
  Input <- tibble(Text = RR[[1]]) %>%
    filter(Text != "") %>% # remove blank entries
    mutate(ID = seq(1:length(Text))) %>% # ID variable
    select(ID, Text) %>% # Right order
    mutate(List_of_Lists =
             str_split(Text, pattern = "(\\t)"))

DATA <- List_to_tibble(Input$List_of_Lists, NAMES=FALSE, PRINT=FALSE)

NAME <- paste0('SAS_', ITEM)

assign(quo_name(enquo(NAME)),
       Input %>%
         cbind(DATA),
       envir=.GlobalEnv)

rm(DATA, RR, Input)
}
```

## Checking

The definitive list of variables is in SAS_TYPE. So we work from there.

Check the names. All of these outputs need to be blank.

```{r}
if (Types$Valid[1]){
Names <- SAS_Type %>%
  full_join(SAS_Label,  by = join_by(Name), keep=TRUE) %>%
  select(starts_with('Name')) %>%
  filter(is.na(Name.x) | (Name.x != Name.y)) ; Names
}

if (Types$Valid[1]){
Names <- SAS_Type %>%
  full_join(SAS_Category,  by = join_by(Name), keep=TRUE) %>%
  select(starts_with('Name')) %>%
  filter(is.na(Name.x) | (Name.x != Name.y)) ; Names
}

if (Types$Valid[3]){
Names <- SAS_Type %>%
  full_join(SAS_Format,  by = join_by(Name), keep=TRUE) %>%
  select(starts_with('Name')) %>%
  filter(is.na(Name.x) | (Name.x != Name.y)) ; Names
}

if (Types$Valid[4]){
Names <- SAS_Type %>%
  full_join(SAS_Label,  by = join_by(Name), keep=TRUE) %>%
  select(starts_with('Name')) %>%
  filter(is.na(Name.x) | (Name.x != Name.y)) ; Names
}

if (Types$Valid[5]){
Names <- SAS_Type %>%
  full_join(SAS_QLU,  by = join_by(Name == SASCode), keep=TRUE) %>%
  select(Name, Question, SASCode)

  Names %>%
    filter(Question != SASCode | Name != SASCode)
  
}
```

```{r Link structure files}
# We use a set of left outer joins
if (Types$Valid[6]){ 
  SAS_Structure <- SAS_Type 
}

if (Types$Valid[3]){ 
  SAS_Structure <- SAS_Structure %>%
    left_join(SAS_Format, by = join_by(Name))
}
if (Types$Valid[4]){ 
  SAS_Structure <- SAS_Structure %>%
    left_join(SAS_QLU, by = join_by(Name == SASCode))
}
if (Types$Valid[5]){ 
  SAS_Structure <- SAS_Structure %>%
    left_join(SAS_Label, by = join_by(Name))
}

# Check this
SAS_Structure %>%
  filter(Label != Description) %>%
  select(ID.x,Name,Question,Description, Label)
# Should be no rows in the output

 
# List of all the SAS datatypes used
SAS_Structure %>%
  select(Type.x, Type.y) %>%
  distinct()

```

```{r Check structure files}
# These ought to be the same
SAS_Structure %>%
  select(Type.x,Format) %>%
  filter(!is.na(Format))
```

# Now we use all of this to read the data properly.

```{r}
# List of all the SAS datatypes used
SAS_Structure %>%
  select(Type.x, Type.y) %>%
  distinct()
# If they start with a $ - they are characters.
# If they start with a number - they are numeric
# If they start with dd - It's a date in the ddmmyy10. format e.g. 12/04/2021

SAS_Types_used <- SAS_Structure %>%
  select(Type.x, Type.y) %>%
  distinct() %>%
  mutate(Type.R = case_when(
    str_detect(Type.x, "^\\$") ~ 'character',
    str_detect(Type.x, "^dd") ~ 'Date',
    str_detect(Type.x, "^-?\\d+\\.?\\d*$") ~ 'numeric',
    .default = 'WARNING - please fix'
  ))

#Now we use this to get the colClasses needed
SAS_Structure <- SAS_Structure %>%
  full_join(SAS_Types_used, by = join_by(Type.x == Type.x, Type.y == Type.y))


# Basic strategy is to use read.delim on the text data
# 
Data <- read.delim(text = SAS_Data$Text,
                   header = FALSE,
                   col.names = SAS_Structure$Name,
                   colClasses = SAS_Structure$Type.R)
str(Data)

# Next we add the variable labels to the data from SAS_Label$Label
# 
# Source : https://strengejacke.github.io/sjlabelled/articles/intro_sjlabelled.html#variable-labels
#

Data %>% var_labels

```
