---
title: "Read SAS directory"
author: "Anthony Staines"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

GERONTE <https://geronteproject.eu/> is a Research and innovation project funded by the European Union within the framework of the H2020 Research and Innovation programme - Transforming Patient Care under grant agreement No. 945218.


```{r setup, include=FALSE}
rm(list = ls())
library(stringr)
library(tidyverse)
library(haven)
library(labelled)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic logic

Go to a directory.
Each contains six files

e.g.

```
GERONTE_FRONE20230407_SAS_Category.txt  GERONTE_FRONE20230407_SAS_Label.txt
GERONTE_FRONE20230407_SAS_Data.txt      GERONTE_FRONE20230407_SAS_QLU.txt
GERONTE_FRONE20230407_SAS_Format.txt    GERONTE_FRONE20230407_SAS_Type.txt
```

The key is the name before the .txt.

```{r Basic lists of files and directories etc}
Directories <- list.files(path=".")
    Directories <-  # Exclude most files
      Directories[
        !str_detect(Directories, '\\.')]
## APN
APN <- list.files(path = Directories[[1]],
                  recursive = TRUE)

##EOS
EOS <- list.files(path = Directories[[2]],
                  recursive = TRUE)

##etc...

## File types - there are six.
Type <- "_([A-z]+)\\."

Types <- str_extract(APN, Type)
    Types <- str_replace(Types,"_SAS_", "")
      Types <- str_replace(Types, "\\.", "")

rm(Type)
```

# Functions

```{r SWAP function}

# Need to swap 1='Yes' to 'Yes'=1

Swap_over_equal_sign <- function(Dataframe, Text) {
  
  Text = Dataframe$Text

  result <- str_match(Text, "(^[\\w]+)(=)(.+)")
  # Three matching groups
  # (1) A word from the beginning
  # (2) A literal = sign
  # (3) The rest of the line to the end

  results <- tibble(V1 = result[,1],
                    V2 = result[,2],
                    V3 = result[,3],
                    V4 = result[,4])
  
  Results <- results %>%
    mutate(V5 = # Put the terminal ; in a separate field
             ifelse(str_detect(V4,';'),';','')) %>%
    mutate(V4 = # Lose the original ;
             str_replace(V4, ';', '')) %>%
    mutate(Swapped = # Glue the pieces together backwards
             paste0(V4,V3,V2)) %>%
    select(Swapped, V1, V2, V3, V4, V5) # Reorder as desired

return(Results)
}
```


```{r List_to_tibble}

#
# From https://stackoverflow.com/questions/42642266/turn-a-list-of-lists-with-unnamed-entries-into-a-data-frame-or-a-tibble
# # Credit to A5C1D2H2I1M1N2O1R2T1
#

List_to_tibble <- function(inlist, PRINT = FALSE) {
  Names <- unlist(inlist[1]) # First row as a character variable

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("List of lists \n")
    print(str(inlist))
    print("\n\n===========================================\n\n")}

  x <- as.data.frame(do.call(rbind, inlist[-1]))

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("as.data.frame(do.call(rbind, inlist[-1])) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}

  x[] <- lapply(x, unlist)

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print(" x[] <- lapply(x, unlist) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}
  
  names(x) <- Names
  x <- tibble(x)

  if(PRINT) {
    print("\n\n===========================================\n\n")
    print("x <- tibble(x) \n")
    print(str(x))
    print("\n\n===========================================\n\n")}

  return(x)
}

```

Work through this one subdirectory and one file at a time

# SAS_Category

These look like this one, and hold value labels.

Value APN_C 1='Yes'
            0='No';
Value TYP1APN 1='Face to face'
              2='Web based'
              3='Phone call';
Value CHA1ACT 1='Yes'
              0='No';

To read these
Read the whole lot in, ignoring carriage returns, so separate only at ; 
Swap equals so 1='Yes' becomes 'Yes'=1 etc...

```{r Read SAS_Category}
RR <- readLines("APN/GERONTE_FRONE20230407_SAS_Category.txt", skipNul = TRUE)
RR <- str_replace_all(RR, "\xff\xfe", "")
str(RR)
```

```{r Process SAS_Category}
Input <- tibble(Text = RR) %>%
  filter(Text != "") %>% # remove blank entries
  mutate(ID = seq(1:length(Text))) %>% # ID variable
  select(ID, Text) %>% # Right order
  mutate(Value = # Create a logical Value to identify lines with variable names
           str_detect(Text, "^Value ")) %>% # The text 'Value ' at the start of a line
  mutate(Text = # Remove the initial Value from those lines
           str_replace(Text, "^Value ", "")) %>%
  mutate(Name = # Create a column with the variable name
           str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the [] bracket
  fill(Name) %>% # Name is only filled for the first code in each variable, so fix this.
  mutate(Text = #Remove the variable name from the Text
           str_replace(Text, "^[\\w]+", "")) %>%
  mutate(Text = str_trim(Text)) # Remove whitespace

rm(RR)

Results <- Input %>% Swap_over_equal_sign(Text) # Text is field with labels in it.

Output <- cbind(Input , Results)

rm(Input, Results)

# Each variable now has several rows. We want a structure where each variable has one row, with the right number of columns.
# 

SAS_CATEGORY <- Output %>%
  group_by(Name) %>%
  summarise(LABEL = toString(Swapped)) %>%
  ungroup()

rm(Output)
```

# SAS_Format

This contains formats for date variables only, for some reason?
Looks like this - 
DT_1APN ddmmyy10.
DT_2APN ddmmyy10.
DT_3APN ddmmyy10.
DT_4APN ddmmyy10.
DT_5APN ddmmyy10.

```{r Read SAS_FORMAT}
RR <- readLines("APN/GERONTE_FRONE20230407_SAS_Format.txt", skipNul = TRUE)
RR <- str_replace_all(RR, "\xff\xfe", "")
str(RR)
```

```{r Process SAS_FORMAT}
Input <- tibble(Text = RR) %>%
  filter(Text != "") %>% # remove blank entries
  mutate(ID = seq(1:length(Text))) %>% # ID variable
  select(ID, Text) %>% # Right order
  mutate(Name = # Create a column with the variable name
           str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
 mutate(Text = #Remove the variable name from the Text
           str_replace(Text, "^[\\w]+", "")) %>%
  mutate(Text = str_trim(Text)) # Remove whitespace

SAS_FORMAT <- Input %>%
  rename(Format = Text) %>%
  select(ID, Name, Format)
```

# SAS_Label

These are variable labels, and look like
PT_CEN="Patient Center derivé"
PT_NUM="Patient Number derivé"
PT_LET="Patient Letter derivé"
APN_C="APN"
DT_1APN="Date APN 1"
TYP1APN="Type APN 1"
CHA1ACT="Interv 1 : Treatment"
CHA1NB="Treatment 1 : Number"
GEN1ACT="Interv 1 : Lifestyle"
GEN1NB="Lifestyle 1 : Number"
CAL1ACT="Interv 1 :Extra HPC"
CAL1NB="Extra HPC 1 : Number"
HPC1ACT="Interv 1 : HPC"
HPC1NB="HPC 1 : Number"
HOS1ACT="Interv 1 : Hospit"


```{r Read SAS_LABEL}
RR <- readLines("APN/GERONTE_FRONE20230407_SAS_Label.txt", encoding = 'UTF-8', skipNul = TRUE)
RR <- str_replace_all(RR, "\xff\xfe", "")
str(RR)
```

```{r Process SAS_LABEL}
Input <- tibble(Text = RR) %>%
  filter(Text != "") %>% # remove blank entries
  mutate(ID = seq(1:length(Text))) %>% # ID variable
  select(ID, Text) %>% # Right order
  mutate(Name = # Create a column with the variable name
           str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
  mutate(Text = #Remove the variable name and the equals sign from the Text
           str_replace(Text, "^[\\w]+=", "")) %>%
  mutate(Text = # Remove the initial an final " from the SAS types
           str_replace_all(Text, "\"", "")) %>%
  mutate(Text = str_trim(Text)) # Remove whitespace

rm(RR)

SAS_LABEL <- Input %>%
  rename(Label = Text) %>%
  select(ID, Name, Label)

rm(Input)
```

# SAS QLU

Looks like this 

Visit/Form/Question	SASCode	Description	Type
APN/APN_CON/PT_CEN	PT_CEN	Patient Center derivé	Integer number
APN/APN_CON/PT_NUM	PT_NUM	Patient Number derivé	Integer number
APN/APN_CON/PT_LET	PT_LET	Patient Letter derivé	Text


```{r Read SAS_QLU}
RR <- readLines("APN/GERONTE_FRONE20230407_SAS_QLU.txt", encoding = 'UTF-8', skipNul = TRUE)
RR <- str_replace_all(RR, "\xff\xfe", "")
str(RR)
```


```{r Process SAS_QLU}
Input <- tibble(Text = RR) %>%
  filter(Text != "") %>% # remove blank entries
  mutate(ID = seq(1:length(Text))) %>% # ID variable
  select(ID, Text) %>% # Right order
  mutate(Text = str_replace(Text, 'Date/Time', 'Date_Time')) %>%
  mutate(List_of_Lists =
           str_split(Text, pattern = "(/|\\t)"))

rm(RR)

QLU <- List_to_tibble(Input$List_of_Lists)
QLU <- QLU %>%
    mutate(Type = str_replace(Type, 'Date_Time', 'Date/Time'))

SAS_QLU <- Input %>%
  slice_head(n = -1) %>% # Remove 'Names' row
  cbind(QLU) %>%
  rename(QLU = Text)

rm(Input)
```

# SAS_Type

Looks like this
Trial :$13.
Site :$8.
Label :$50.
PersonId :5.
VisCycle :5.
FrmCycle :5.
RepeatNo :5.
PT_CEN :2.
PT_NUM :3.
PT_LET :$3.
APN_C :2.
DT_1APN :ddmmyy10.

```{r Read SAS_Type}
RR <- readLines("APN/GERONTE_FRONE20230407_SAS_Type.txt", skipNul = TRUE)
RR <- str_replace_all(RR, "\xff\xfe", "")
str(RR)
```

```{r Process SAS_TYPE}
Input <- tibble(Text = RR) %>%
  filter(Text != "") %>% # remove blank entries
  mutate(ID = seq(1:length(Text))) %>% # ID variable
  select(ID, Text) %>% # Right order
  mutate(Name = # Create a column with the variable name
           str_extract(Text, "^[\\w]+")) %>% ## + has to be outside the  [] bracket
  mutate(Text = #Remove the variable name from the Text
           str_replace(Text, "^[\\w]+", "")) %>%
  mutate(Text = str_trim(Text)) %>% # Remove whitespace
  mutate(Text = # Remove the initial : from the SAS types
           str_replace(Text, "^:", "")) %>%
  mutate(Text = str_trim(Text)) # Remove whitespace

rm(RR)

SAS_TYPE <- Input %>%
  rename(Type = Text) %>%
  select(ID, Name, Type)

rm(Input)
```


